---
title: "CDC Fluoridation googleVis Chart"
author: "Brian High"
date: "05/21/2015"
output: 
    html_document:
        keep_md: yes
---

## Setup

Load the required R packages.

```{r, echo=TRUE, message=FALSE}
for (pkg in c("XML", "googleVis")) {
    if (! suppressWarnings(require(pkg, character.only=TRUE)) ) {
        install.packages(pkg, repos="http://cran.fhcrc.org", dependencies=TRUE)
        if (! suppressWarnings(require(pkg, character.only=TRUE)) ) {
            stop(paste0(c("Can't load package: ", pkg, "!"), collapse = ""))
        }
    }
}
```

## Get the data

Load the CSV data file if you have it, otherwise create the CSV by scraping the 
fluoridation table from the CDC website.

```{r}
cdcfile <- "data/uw_cdc_dw_fluoridation.csv"
if (file.exists(cdcfile)) {
    # Read data from CSV
    cdcfl <- read.csv(file = cdcfile, header = TRUE)
} else {
    # Scrape from CDC web page and save as CSV for later
    library(XML)
    cdcflweb<-readHTMLTable("http://www.cdc.gov/fluoridation/statistics/2012stats.htm")
    cdcfl <- cdcflweb[[2]][3:53, c(1,4)]
    names(cdcfl) <- c("State", "Percent")
    cdcfl$State <- sapply(cdcfl$State, function (x) gsub("f[,gh]*$", "", x))
    write.csv(cdcfl, file=cdcfile, row.names=FALSE)
}
```

## Clean the data

Remove the % sign from Percent and convert to numeric type.

```{r}
cdcfl$Percent <- as.numeric(sub("%", "", cdcfl$Percent))
```

## Create the gvisGeoChart

```{r ComboExample, results='asis', tidy=FALSE}
# Set options
op <- options(gvis.plot.tag='chart')

# Create gvisGeoChart
suppressPackageStartupMessages(library(googleVis))
G3 <- gvisGeoChart(cdcfl, "State", "Percent",
                    options=list(region="US", displayMode="regions",
                                 resolution="provinces",
                                 width=600, height=400))
plot(G3)

# Set options back to original options
options(op)
```

## Example Web Page

The HTML output from running `knitr` on the Rmd file (the original version of 
this document) is a stand-alone web page and can be found on 
[Brian High's UW website](http://staff.washington.edu/high/fluoridation/cdc-fl-example.html)